---
title: "Generalized linear model"
author: Peter Smits
output: beamer_presentation
---
```{r constants, echo=FALSE, message=FALSE}
# set up the R environment

if(!('pacman' %in% rownames(installed.packages()))) {
  install.packages('pacman')
}
library(pacman)

p_load(readr, janitor, here)
p_load(tidyverse, broom, purrr, ggplot2)

if(!file.exists('.here')) {
  set_here(path = getwd())
}
```


# Review of Linear Regression



## Definition of linear regression

> ... a method that summarizes how the average values of a numerical *outcome* variable vary over subpopulations defined by linear functions of *predictors*. [...] Regression can be used to predict an outcome given a lienar function of these predictors, and regression coefficients can be thought of as comparisons across predicted values or as comparisons among averages in the data.

\begin{small}
  Gelman and Hill, 2007, p.31
\end{small}



## Compacy written form

\begin{align*}
y &\sim \mathcal{N}(\mu, \sigma) \\
\mu &= X \beta
\end{align*}

\(y \in \mathbb{R}\), \(\mu \in \mathbb{R}\), \(\sigma \in \mathbb{R}^{+}\), \(\beta_{k} \in \mathbb{R} \text{for } k = 1, ..., K\).

\(y\) is a length \(N\) vector of observations. \(X\) is a \(N \times K\) matrix of covariates (and a column of 1s). \(\beta\) is a length \(K\) vector of regression coefficients (including intercept).



## Interpreting regression parameters
  
Coeffcient \(\beta\) is the expected difference in \(y\) between two observations that differ by 1 in a single predictor.



## Fitting and inspecting a regression model

```{r ex_linear}


```

